---
title: "**PVWAT NOTEBOOK: DUST-TO-SOILING SIMULATION**"
author: "Amy Tamunoibinyemiem Banigo, Louise Crochemore"
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, error=FALSE, fig.cap = " ", message=FALSE, warning=FALSE)
```

------------------------------------------------------------------------

# PVWAT NOTEBOOK

## DUST-TO-SOILING SIMULATION

------------------------------------------------------------------------

### **General Info**

PVWAT (Banigo et al.) allows one to simulate time series of the dust deposition flux on panels, the amount of dust accumulated on panels, and the associated soiling losses, from time series of meteorological variables (precipitation intensities, wind speed) and particulate matter concentrations for different particle classes. In the following, the model is applied for a single location representing a hypothetical solar farm with identical PV panels.

PVWAT consists of three sub-models: a deposition flux model, a dust accumulation model and a soiling loss model. These were specifically chosen for their adaptability to data-scarce regions.

This notebook contains the functions needed to run PVWAT and describes how to simulate dust influx, dust accumulation and soiling losses from meteorological and particulate matter time series at a given location.

------------------------------------------------------------------------

### **1. Required Inputs**

#### Input Time Series

Input data files must contain for each location a list containing 3-hourly time series with dates, wind speed, precipitation, particulate matter concentration for each class of particle. It also contains the characteristic radius $A$ and roughness length $z_0$ that both depend on the land use category at this location.

```{r}
# Set directory for input time series
input_dir <- ""

# Set name of .RDS file containing input time series for one location
point_data_file <- "point_data.rds"

# Set directory for output time series
output_dir <- input_dir
  
```

#### **Location name, Coordinates and Dry season dates**

```{r}
# Data frame containing the main characteristics about the location where simulation runs
PVWAT_point <- data.frame(
  name             = "Location_1",
  lon              = 0,
  lat              = 0,
  stringsAsFactors = FALSE
)

cat("\nPVWAT point created...\n")
cat(" Site: ",PVWAT_point$name, 
    "\n Longitude:",PVWAT_point$lon, 
    "\n Latitude:",PVWAT_point$lat, "\n")

```

#### **Particulate Matter Concentration Classes**

```{r}
# PM size classes based on PM data times series used (here, CAMS)
# Name and labels and mean diameter to consider for each class
PM_CLASSES <- data.frame(
  name = c("PM1", "PM2.5-PM1", "PM10-PM2.5", "PM20-PM10"),
  label = c("PM1 (dp=1µm)", "PM2.5-PM1 (dp=1.75µm)", 
            "PM10-PM2.5 (dp=8µm)", "PM20-PM10 (dp=15µm)"),
  dp = c(1e-6, 1.75e-6, 8e-6, 15e-6),
  stringsAsFactors = FALSE
)
```

------------------------------------------------------------------------

### 2. Import time series and parameters

#### **Load atmospheric time series**

```{r}
# Append input directory and file name 
input_file = file.path(input_dir,point_data_file)
 
# Check that file exists
if (!file.exists(input_file)) {
  stop("point_data.rds not found. 
       Make sure you specified an input directory and placed the file there.")
}

# Import file
point_data <- readRDS(input_file)
```

#### **Set physical parameters**

```{r}
kappa   <- 0.41          # von Karman constant
g       <- 9.81          # acceleration due to gravity
rhop    <- 2200          # density of aerosol particles
mu      <- 1.849e-5      # dynamic viscosity of air at 25 degrees C
v       <- 1.562e-5      # kinematic viscosity of air at 25 degrees C
k_boltz <- 1.38065e-23   # Boltzmann constant
h_Vd    <- 10            # reference height set to 10 m for ERA5 data
```

### 3. Load PVWAT model functions

#### **Cunningham Correction Function**

The Cunningham correction factor is used to account for non--continuum effects when calculating the drag force on small particles.

```{r}
cunningham_correction <- function(dp) {

  lambda <- 6.73e-8
  dp_cutoff <- 2.5e-6

  A1 <- 1.257
  A2 <- 0.4
  A3 <- 1.1

  Cc <- ifelse(
    dp >= dp_cutoff,
    1,
    1 + (lambda / dp) *
      (A1 + A2 * exp(-A3 * dp / lambda))
  )

  return(Cc)
}
```

#### **Dust Deposition Velocity Model Function**

The deposition velocity of a particle $V_d(t)$ is estimated by considering gravitational settling, resistance to atmospheric turbulence, and surface resistance and is expressed as:

$V_d(t) = \frac{1}{R_a(t) + R_s(t)} + V_s(t) \cdot \cos\theta$

Where $R_a(t)$ and $R_s(t)$ are the aerodynamic and the surface resistance terms (s/m), $V_s(t)$ is the gravitational settling velocity (m/s) and $\theta$ is the tilt angle of the panel, here assumed equal to the latitude, the suggested optimum inclination for solar panels (Hosseini et al., 2023 ; Sharma et al., 2023).

```{r}
DepositionVelocityModel <- function(W, Tilt, dp, z0, A, gam, epsilon = 3.0, BSct = 1.0) {
  
  # roughness length
  z0_safe <- pmin(z0, h_Vd * 0.5)
  
  # Cunningham correction factor
  Cc <- cunningham_correction(dp)
  
  # Friction velocity
  u <- abs((kappa * W) / log(h_Vd / z0_safe))
  
  # Schmidt number
  Sc <- v / ((k_boltz * 298.15 * Cc) / (3 * pi * mu * dp))
  
  # Gravitational settling velocity
  Vsf <- (g * rhop * (dp^2) * Cc) / (18 * mu)
  
  # Aerodynamic resistance
  Rat <- (BSct / (kappa * u)) * log(h_Vd / z0_safe)
  
  # Surface resistance
  St <- ((u^2) * Vsf )/ (v * g)
  Eb <- Sc^(-gam)
  Eim <- St^2 / (400 + St^2)
  Ein <- ifelse(is.na(A), 0, pmin(1, 0.5 * (dp / A)^2))
  R1 <- exp(-sqrt(St))
  Rb <- 1 / (u * epsilon * (Eb + Eim + Ein) * R1)
  
  # Deposition velocity
  Vd <- (1 / (Rat + Rb) + Vsf * cos(Tilt * pi / 180)) * 100
  
  return(Vd)
}
```

#### **Dust Accumulation Model Function**

Dust accumulation model relies on the mass balance equation for dust:

$\frac{d \omega(t)}{dt} = i(t) - o_w (t)+o_p (t)$

with $\omega(t)$ the amount of dust accumulated on the panels per unit area (g/m$^{2}$), $i(t)$ the dust deposition flux to the panel, $o_w(t)$ the removal flux due to wind and $o_p(t)$ the removal flux due to precipitation.

Following Besson et al (2017), precipitation is assumed to remove all accumulated dust once precipitation intensity exceeds a threshold $P_0$. For wind removal, we introduce a bulk removal model that assumes that the removal is proportional to the mass of dust accumulated on the panels at time $t$:

$o_{w}(t) = \frac{\omega(t)}{\frac{k}{u^2}}$

with $k$ the wind removal constant and $u$ the wind speed.

```{r}
AccumulationModel <- function(precip, flux, wind, h, Tr, k_base = 700) {
  M <- rep(0, length(flux))
  
  # Bulk wind removal term
  k <- k_base / wind^2
  
  # Loop on time steps
  for (i in 1:length(flux)) {
    
    # Display state
    if(i%% 1000 ==0 || i==1) cat('      ',i,'/',length(flux),'\n')
    
    # If missing influx of particulate matter or wind speed
    if (is.na(flux[i]) || is.na(wind[i])) {
      M[i] <- NA
      next
    }
    # Initialisation and Reinitialisation
    if (i == 1 || is.na(M[i-1])) {
      M[i] <- flux[i]
      next
    }
    # Wind or precipitation removal
    # Conditions for wind removal but no precipitation removal
    condition1 <- (wind[i] >= 0.000001 && precip[i] < Tr)
    condition2 <- (wind[i] >= 0.000001 && is.na(precip[i]))
    if (condition1 || condition2) {
      M[i] <- M[i-1] * exp(-h/(24 * k[i])) + k[i] * 24/h * 
        ((flux[i] + flux[i-1]) / 2) * (1 - exp(-h/(24 * k[i])))
    }
  }
  return(M)
}
```

#### **Soiling Model Function**

Soiling losses are calculated from the amount of dust accumulated on panels. They represent the reduction in PV production efficiency with respect to the efficiency of a clean panel. The function used here is generic, parsimonious and flexible (cf. Banigo et al.):

$SR(t) = (100 - p_1) + p_1 \cdot \exp(p_2 \cdot \omega(t))$

```{r}
SoilingModel <- function(X, p1 = 67.72, p2 = -0.29) {
  
  return((100 - p1) + p1 * exp(p2 * X))

}
```

------------------------------------------------------------------------

### 4. Load PVWAT simulation function

This function runs the entire PVWAT modelling chain, calling all models : dust deposition, dust accumulation and soiling ratio.

```{r}
Compute_PVWAT <- function(pm, precip, wind, dates, tilt_angle, z0, A, gamma, Tr, h) {
  
  cat("\n#####################################################\n")
  cat("### SIMULATION START\n")
  cat("#####################################################\n")
  
  # FORCE TO VECTORS
  wind <- as.vector(wind)
  precip <- as.vector(precip)
  
  # Calculate flux (g/m2/hr) with PM and wind
  cat("     Calculating dust influx from PM and wind...\n")
  flux <- rep(0, length(wind))
  # Cunningham correction table
  dps_tab <- c(1e-6, 2.5e-6, 1e-5, 2e-5)
  CCs_tab <- c(1.164, 1.073667, 1.016, 1.008)
  # Calculate flux for each class of particulate matter
  for (i in 1:nrow(PM_CLASSES)) {
    dp_val_i <- PM_CLASSES$dp[i]
    # ALSO force PM data to vector
    pm_ts_i <- as.vector(pm[[PM_CLASSES$name[i]]])
    Vd_ts <- DepositionVelocityModel(wind, tilt_angle, dp_val_i, z0, A, gamma)
    flux <- flux + (pm_ts_i * Vd_ts * 36000)
  }
  
  # Calculate dust accumulation from dust influx, precipitation and wind
  cat("     Calculating dust accumulation...\n")
  k_base <- 700
  acc <- AccumulationModel(precip, flux, wind, h, Tr, k_base)
  
  # Calculate soiling loss from dust accumulation
  cat("     Calculating soiling loss...\n")
  soiling <- SoilingModel(acc)
  
  cat("\n#####################################################\n")
  cat("### SIMULATION COMPLETE\n")
  cat("#####################################################\n")
  
  return(data.frame(
    dates = dates,
    dust_influx = flux,
    dust_acc = acc,
    soiling_ratio = soiling
  ))
}

```

------------------------------------------------------------------------

### **5. Run PVWAT**

Here, input time series and parameters are imported, checked and used as input to the PVWAT modelling chain. Resulting simulations of dust influx, dust accumulation and soiling ratio are then checked and exported.

#### PVWAT simulation

```{r}
h_timestep <- 3
Tr_ref <- 1 / 8

# Set panel tilt angle to latitude
Tilt_val <- PVWAT_point$lat

# Location-dependent input dates, time series and parameters
dates        <- point_data$dates
wind_ts      <- point_data$wind
precip_ts    <- point_data$precip
pm_ts        <- point_data$pm_data_list
z0_val       <- point_data$z0
A_val        <- point_data$A
gamma_val    <- point_data$gamma

# Removing negative wind values
wind_ts[wind_ts   <= 0] <- NA
precip_ts[precip_ts <= 0] <- NA

cat(sprintf(
  "Parameters: z0=%.6f, A=%s, gamma=%s\n",
  z0_val,
  ifelse(is.na(A_val), "NA", sprintf("%.6f", A_val)),
  ifelse(is.na(gamma_val), "NA", sprintf("%.6f", gamma_val))
))

cat("\nRunning PVWAT...\n\n")

PVWAT_outputs <- Compute_PVWAT(pm = pm_ts, precip = precip_ts, wind = wind_ts, 
  dates = dates, tilt_angle = Tilt_val, z0 = z0_val, A = A_val,
  gamma = gamma_val, Tr = Tr_ref, h = h_timestep)


```

#### Check PVWAT simulations

```{r}
# Check outputs
print(head(PVWAT_outputs))

# Plot output time series for max 3 years
par(mfrow=c(3,1),las=2)
N = 3*365*24/h_timestep
if(nrow(PVWAT_outputs) < N) N = nrow(PVWAT_outputs)
par(mar=c(1,4,4,2)) ; plot(PVWAT_outputs$dust_influx[1:N],type='l',xaxt='n',
                           ylab='Dust influx (g/m2/h)')
abline(v=which(point_data$precip >= Tr_ref,1,0),col='blue',lwd=1)
par(mar=c(2.5,4,2.5,2)) ; plot(PVWAT_outputs$dust_acc[1:N],type='l',xaxt='n',
                               ylab='Dust accumulation (g/m2)')
par(mar=c(4,4,1,2)) ; plot(PVWAT_outputs$soiling_ratio[1:N],type='l',xaxt='n',
                           ylab='Soiling ratio (%)',xlab='')
at = which(substr(PVWAT_outputs$dates,6,13)=='01-01')
axis(1,at=at, labels = substr(PVWAT_outputs$dates[at],1,4))
```

#### Export PVWAT simulation

```{r}
# Save results
output_file <- file.path(output_dir,"PVWAT_results.csv")
write.csv(PVWAT_outputs, output_file, row.names = FALSE,quote = F)
```
